TEMPERATURES AND METALLICITIES FOR 10,000 M DWARFS\\ IN THE \apogee\ SURVEY

### 2019-06-10 paper/draft.tex

Low-mass stars (with masses $M_{*} < 0.7\,\msun$ and atmospheric temperatures $T_{\rm eff} < 4000\,$K) are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010}. With nuclear fusion timescales of $\tau>10^{11}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010}.

Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity offset detections \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations \citep{Nutzman:2008}, and shorter orbital periods allow for new planet discovery in less observation time than their higher mass counterparts. For these reasons, M dwarfs are the primary candidates for exoplanet detection searches (e.g. the NASA \kepler\ and Transiting Exoplanet Survey Satellite (\tess) missions), making detailed and precise knowledge of M dwarf chemical compositions key to constraining the formation scenarios and atmospheric conditions of potentially habitable exoplanets.

Advances in instrumentation with the lauch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources) and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III/IV mission \citep{Eisenstein:2011,Blanton:2017} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.

While elemental abundance measurements from high resolution spectra of F, G, and K stars have achieved extremely high precisions (down to errors in the range of $0.01-0.03$\,dex; \citealt{Nissen2018}) enabled by improvements in atmospheric models towards realistic 3D local thermodynamic equilibrium (LTE) assumptions \citep{Asplund2005}, and differential abundance techniques using equivalent widths \citep{Bedell2014},
the determination of precise metallicities for M dwarfs has remained a long standing challenge due to the formation of complex diatomic and triatomic molecular bands at M dwarf temperature regimes, particularly TiO and VO which dominate the optical, H$_2$O and CO which dominate the infrared, and hydride bands (FeH, CaH, CrH, MgH, etc.) which form at late spectral types \citep{Allard1997}.
In particular, atmospheric models fail to reproduce spectral features due to incomplete line lists and opacities, and the presence of millions of weak, blended transitions and lack of continuum in a spectrum make it difficult to deconvolve individual features and extract line strengths from equivalent widths,
creating fundamental limitations in efforts towards exploiting the information content available out of high resolution datasets using traditional modeling and empirical methods.

A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically associated wide binaries are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g. \citealt{Terrien:2012}; \citealt{Rojas-Ayala:2012}; \citealt{Newton:2014}).
However these metallicity calibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant calibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.

In this work we build a data-driven model for M dwarf spectra with \thecannon\ \citep{Ness:2015,Casey:2016,Ho:2017a}.
\thecannon\ is a pixel-by-pixel regression for spectral data given a training
set of stars with reliable (assumed) accurate labels.
The model can both predict a spectrum given labels, and also predict labels given a spectrum.
Thus it is a generative model for stellar spectra that can be used for parameter inference.

The data-driven approach of \thecannon\ is ideal in certain cases:
if stellar labels are only known for a small number of stars
but there are spectra taken for many more;
if it is computationally expensive to obtain labels for a star, but
there are many stars that need labels;
if there are spectral models that work in one wavelength range or resolution but not in another.
The context of this paper is the last:
there are precise calibrations for metallicities of M dwarf stars in the optical, which have been
used successfully, but (as we note above) there isn't as much success with such modeling in the infrared, and not at \apogee\ resolution.
\thecannon\ bridges this gap: it doesn't require using specific lines; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.

>>>>

Low-mass stars (with masses $M_{*} < 0.7\,\msun$ and atmospheric temperatures $T_{\rm eff} < 4000\,$K) are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010}. With nuclear fusion timescales of $\tau>10^{11}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010,Woolf:2012}.

Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity offset detections \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations \citep{Nutzman:2008}, and shorter orbital periods (for a fixed stellar insulation flux) allow for new planet discovery in less observation time than their higher mass counterparts. For these reasons, M dwarfs are primary candidates for exoplanet searches, including by the NASA \kepler\ \citep[e.g.,][]{Dressing2015} and Transiting Exoplanet Survey Satellite (\tess) \citep[e.g.,][]{Muirhead:2018} missions. As a result, detailed and precise knowledge of M dwarf chemical compositions has become key to constraining the the properties, formation scenarios, and atmospheric conditions of potentially habitable exoplanets observable with {\it JWST}.

Advances in instrumentation with the launch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources), enabling studies of M dwarf abundances on both exoplanet and Galactic scales.
The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}. \apogee\ pipeline measurements of $\teff$ and $\feh$ \citep{Perez:2016} for M dwarfs have only been determined to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.

Elemental abundance measurements from high resolution spectra of F, G, and K stars have achieved extremely high precision (down to $0.01-0.03$\,dex; \citealt{Nissen2018}) enabled by improvements in atmospheric models, including realistic 3D local thermodynamic equilibrium (LTE) assumptions \citep{Asplund2005}, and differential abundance techniques using equivalent widths \citep{Bedell2014}. However, the determination of precise metallicities for M dwarfs has remained a long-standing challenge due to the formation of complex diatomic and triatomic molecular bands at M dwarf temperature regimes, particularly TiO and VO dominating the optical spectrum, H$_2$O and CO for the infrared, and hydride bands (FeH, CaH, CrH, MgH, etc.) present in the spectra of the latest spectral types \citep{Allard1997}.
In particular, atmospheric models fail to reproduce spectral features \citep[e.g.,][]{Mann:2013c} due to incomplete line lists and opacities, and the presence of millions of weak, blended transitions, and the lack of continuum in a spectrum, all make it difficult to deconvolve individual features and extract line strengths from equivalent widths. The combination of these effects creates fundamental limitations to explore the information content available out of high resolution spectra using a traditional comparison to atmosphere models.

A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically associated wide binaries are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the spectra of the M dwarf (e.g. \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}). However these metallicity calibrations are inaccessible for analysis with \apogee, as the wavelength coverage does not include important calibration indices in the $K$-band or include important information about the spectral shape needed to determine \teff. Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that would fail to utilize the additional spectral information provided by \apogee's resolution.

%the relevant calibration indices used by existing studies (such as Ca~I or K~I lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis. 
%AWM: I'm not really sure this is true. The major lines I and Babs use are in the K-band, very few studies use optical bands. Ryan includes ones in the H-band, some of which are in Apogee. I think the bigger point is that this fails to utilize the high-resolution information of the spectra, and relies instead of temperature calibration from spectral shape, which is impossible at this resolution.

%%AWM: would it be useful to have some comment on data-driven models in general? Their virtues should probably be explained before invoking them

In this work we build a data-driven model for M dwarf spectra with \thecannon\ (\citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}, \citealt{Behmard2019}).
\thecannon\ is a pixel-by-pixel regression for spectral data given a training set of stars with reliable (assumed) accurate labels. The model can both predict a spectrum given labels, and also predict labels given a spectrum. Thus it is a generative model for stellar spectra that can be used for parameter inference.
%AWM: not an expert here, so I won't add anything, but this probably needs more information!

The data-driven approach of \thecannon\ is ideal in certain cases:
if stellar labels are only known for a small number of stars
but there are spectra taken for many more;
if it is computationally expensive to obtain labels for a star, but
there are many stars that need labels;
if there are spectral models that work in one wavelength range or resolution but not in another.
The context of this paper is the last:
there are precise calibrations for metallicities of M dwarf stars in the optical, which have been
used successfully, but (as we note above) there isn't as much success with such modeling in the infrared, and not at \apogee\ resolution.
\thecannon\ bridges this gap: it doesn't require using specific lines; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
%AWM: again, the metallicity metrics Babs/Ryan/Myself use are all in the NIR, and they aren't based on models. The temperature tools use models, but they are calibrated against LBOI measurements. I need to come back to editing later, but this paragraph probably needs to be tweaked a bit.
